A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-12-21 inst/manuscript/manuscript.Rmd
minor updates

---
title: A comparison of design-based and model-based approaches for finite population spatial data.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  1.   The design-based and model-based approaches to frequentist statistical inference lie on fundamentally different foundations. In the design-based approach, inference depends on random sampling. In the model-based approach, inference depends on distributional assumptions. We compare the approaches for finite population spatial data.
  2. We provide relevant background for the design-based and model-based approaches and then study their performance using simulations and an analysis of real mercury concentration data. In the simulations, a variety of sample sizes, location layouts, dependence structures, and response types are considered. In the simualtions and real data analysis, the population mean is the parameter of interest and performance is measured using statistics like bias, squared error, and interval coverage.
  3. When studying the simulations and mercury concentration data, we found that regardless of the strength of spatial dependence in the data, sampling plans that incorporate spatial locations (spatially balanced samples) generally outperform sampling plans that ignore spatial locations (non-spatially balanced samples). We also found that model-based analyses tend to outperform design-based analyses, even when the data are skewed (and by consequence, the model-based distributional assumptions violated). The performance gap between the analysis approaches is small when spatially balanced samples are used but large when non-spatially balanced samples are used. This suggests that the sampling choice (spatially balanced samples versus non-spatially balanced samples) is most important when using a design-based analysis.
  4. There are many benefits and drawbacks practitioners must consider when choosing between design-based and model-based approaches for finite population spatial data. We provide relevant background contextualizing each approach and study their properties in a variety of scenarios, making recommendations for use based on the practitioner's goals.
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \usepackage{caption}
  \usepackage{subcaption}
  \usepackage{setspace}
  \doublespacing
output: rticles::elsevier_article
---

@@

There are two general approaches for using data to make frequentist statistical inferences about a population: design-based and model-based. When data cannot be collected for all units in a population (i.e., population units), data are collected on a subset of the population units. This subset is called a sample. In the design-based approach, inferences about the underlying population are informed via a probabilistic process assigning some population units to be part of the sample. Alternatively, in the model-based approach, inferences are made from specific assumptions about the underlying process generating the data. Each paradigm has a deep historical context [@sterba2009alternative] and its own set of benefits and drawbacks [@hansen1983evaluation].

@@

Though the design-based and model-based approaches apply to statistical inference in a broad sense, we focus on comparing these approaches for spatial data. We define spatial data as data that incorporates the specific locations of the population units into either the design or estimation process. @de1990model give an early comparison of design-based and model-based approaches for spatial data, quashing the belief that design-based approaches could not be used for spatially correlated data. Since then, there have been several general comparisons between design-based and model-based approaches for spatial data [@brus1997random; @verhoef2002sampling; @verhoef2008spatial; @wang2012review; @brus2021statistical]. @cooper2006sampling reviews the two approaches in an ecological context before introducing a "model-assisted" variance estimator that combines aspects from each approach. In addition to @cooper2006sampling, there has been substantial research and development into estimators that use both design and model-based principles (see e.g., @sterba2009alternative, @cicchitelli2012model, @chan2020bayesian for a Bayesian approach).

<!-- Paragraph about the contribution of the manuscript and outlining the rest -->
Certainly comparisons between design-based and model-based approaches to spatial data have been studied. But no numerical comparison has been made between design-based approaches incorporating spatial information and design-based approaches. In this manuscript, we compare design-based approaches incorporating spatial information to model-based approaches for spatial data. We focus on finite populations, but these comparisons generalize to infinite populations as well. A finite population contains a finite number of population units; an example is lakes (treated as a whole with the lake centroid representing location) in the contiguous United States. An infinite population contains an infinite number of population units; an example is locations within a single lake.

The rest of the manuscript is organized as follows. In Section \ref{sec:background}, we introduce and provide relevant background for the design-based and model-based approaches to finite population spatial data. In Section \ref{sec:mm}, we describe how we compare performance of the approaches wiht a simulation study and an analysis of real data that contains mercury concentration in lakes from the contiguous United States. In Section \ref{sec:results}, we present results from the simulation study and the analysis of mercury concentrations. And in Section \ref{sec:discussion}, we end with a discussion and provide directions for future research.

@@

The design-based approach assumes the population is fixed. Randomness is incorporated via the selection of units in a sampling frame. A sampling frame is the set of all units available to be sampled. Units from the sampling frame are selected as part of the sample according to a sampling design, which assigns a positive probability of inclusion (inclusion probability) to each unit from the sampling frame. Some examples of commonly used sampling designs include simple random sampling, stratified random sampling, and cluster sampling. When sampling designs incorporate spatial locations into sampling, we call the resulting samples "spatially balanced." One approach to selecting spatially balanced samples is the Generalized Random Tessellation Stratified (GRTS) algorithm [@stevens2004spatially], which we discuss in more detail in Section \ref{subsec:spb_design}. When sampling designs do not incorporate spatial locations into sampling, we call the resulting samples "non-spatially balanced."

@@

The model-based approach assumes the data are a random realization of a data-generating stochastic process. Randomness is incorporated through distributional assumptions on this process. Strictly speaking, randomness need not be incorporated through random sampling, though @diggle2010geostatistical warn against preferential sampling. Preferential sampling occurs when the process generating the data locations and the process being modeled are not independent of one another. To guard against preferential sampling, model-based approaches often still implement some form of random sampling.

@@

We previously mentioned that the design-based approach can be used to select spatially balanced samples (samples that incorporate spatial locations of the population units and are "well-spread" is space). Spatially balanced samples are useful because parameter estimates from these samples tend to vary less than parameter estimates from samples that are not spatially balanced [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @wang2013design; @benedetti2017spatiallyreview]. The first spatially balanced sampling algorithm seeing widespread use is the Generalized Random Tessellation Stratified (GRTS) algorithm [@stevens2004spatially]. To quantify the spatial balance of a sample, @stevens2004spatially proposed loss metrics based on Voronoi polygons (Dirichlet Tessellations). After the GRTS algorithm was developed, several other spatially balanced sampling algorithms emerged, such as the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance Sampling [@benedetti2017spatially], and Halton Iterative Partitioning Sampling [@robertson2018halton]. In this manuscript, we select spatially balanced samples using the Generalized Random Tessellation Stratified (GRTS) algorithm because it has several attractive properties. More specifically, the GRTS algorithm accommodates finite and infinite sampling frames, equal, unequal, and proportional (to size) inclusion probabilities, legacy (historical) sampling [@foster2017spatially], a minimum distance between units in a sample, and replacement units (replacement units are population units that can be sampled when a population unit originally selected can no longer be sampled). The GRTS algorithm selects samples by utilizing a particular mapping between two-dimensional and one-dimensional space that preserves proximity relationships. Via this mapping, units in two-dimensional space are partitioned using a hierarchical address. This hierarchical address is used to map population units to a one-dimensional line. On the one dimensional line, each population unit's line length equals its inclusion probability. Then, a systematic sample of population units is selected on the line, yielding desired sample. @stevens2004spatially provides more technical details.

After selecting a sample and collecting data, unbiased estimates of population means and totals can be obtained using the Horvitz-Thompson estimator [@horvitz1952generalization]. If $\tau$ is a population total, the Horvitz-Thompson estimate of $\tau$, denoted by $\hat{\tau}_{ht}$, is is given by
\begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align}
where $Z_i$ is the value of the $i$th population unit in the sample and $\pi_i$ is the inclusion probability of the $i$th population unit in the sample. An estimate of the population mean is obtained by dividing $\hat{\tau}_{ht}$ by $N$, the number of population units.

@@

Finite Population Block Kriging (FPBK) is a model-based approach that expands the geostatistical Kriging framework to the finite population setting [@verhoef2008spatial]. Instead of developing inference based on a specific sampling design, we assume the data are generated by a spatial stochastic process. We summarize some of the basic principles of FBPK next (for more technical details, see @verhoef2008spatial) Let ${\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}$ be an $N \times 1$ response vector at locations $s_1$, $s_2$, . . . , $s_N$ that can be measured at the $N$ population units. Suppose we want to use a sample to predict some linear function of the response variable, $f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}$, where $\mathbf{b}^\prime$ is a $1 \times N$ vector of weights (e.g, the population mean is represented by a weights vector whose elements all equal one). Denoting quantities that are part of the sampled population units with a subscript \emph{s} and quantities that are part of the unsampled population units with subscript \emph{u}, let

@@

FBPK assumes $\bm{\delta}$ in Equation$~$\ref{equation:Zmarginal} has mean-zero and a spatial correlation structure that can be modeled using a covariance function. This covariance function is commonly assumed to be non-negative (between zero and one), second-order stationary (depending only on the distance between population units), isotropic (independent of direction), and decay with distance between population units [@cressie1993statistics]. Henceforth, it is implied that we have made these same assumptions regarding $\bm{\delta}$, though @chiles1999geostatistics, pp. 80-93 discuss covariance functions that are not second-order stationary, not isotropic, or both. A variety of flexible covariance functions can be used to model $\bm{\delta}$ [@cressie1993statistics]; one example is the exponential covariance function (for a thorough list of spatial covariance functions, see @cressie1993statistics. The $i,j$th element of the exponential covariance matrix, $\cov(\bm{\delta})$, is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align}
where $\sigma^2_{1}$ is the variance parameter quantifying the variability that is dependent (coarse-scale), $\sigma^2_{2}$ is the variance parameter quantifying the variability that is independent (fine-scale), $\phi$ is the range parameter measuring the distance-decay rate of the covariance, and $h_{i,j}$ is the Euclidean distance between population units $i$ and $j$. The proportion of variability attributable to dependent random error is $\sigma^2_{1} / (\sigma^2_{1} + \sigma^2_{2})$. Similarly, the proportion of variability attributable to independent random error is $\sigma^2_{2} / (\sigma^2_{1} + \sigma^2_{2})$. Finally we note that $\sigma^2_{1}$ and $\sigma^2_{2}$ are often called the partial sill and nugget, respectively.

@@

Other approaches, such as k-nearest-neighbors [@fix1989discriminatory; @ver2013comparison], random forests [@breiman2001random], Bayesian models [@chan2020bayesian], among others, could also be used to obtain predictions for a mean or total from spatially correlated responses of a finite population. Compared to the k-nearest-neighbors and random forest approach, we prefer FBPK because it is model-based and relies on theoretically-based variance estimators leveraging the model's spatial covariance structure, whereas k-nearest-neighbors and random forests use ad-hoc variance estimators [@ver2013comparison]. Additionally, @ver2013comparison studied compared FBPK, k-nearest-neighbors, and random forests in a variety of spatial data contexts, and FBPK tended to perform best. Compared to the Bayesian approach, we prefer FPBK mostly because it is much more computationally efficient.

@@

We used a simulation study to investigate performance of four sampling-analysis combinations: IRS with a design-based analysis, called "IRS-Design"; IRS with a model-based analysis, called "IRS-Model"; GRTS sampling with a design-based analysis, called "GRTS-Design"; GRTS sampling with a model-based analysis, called "GRTS-Model". These combinations are also provided in Table \ref{tab:designanalysis}.

@@

Performance for the four sampling-analysis combinations was evaluated in 36 different simulation scenarios. The 36 scenarios resulted from the crossing of three sample sizes, two location layouts, two response types, and three proportions of dependent random error. The three sample sizes ($n$) were $n = 50, n = 100,$ and $n = 200$. Samples were always selected from a population size ($N$) of $N = 900$. The two location layouts (of the population units) were random and gridded. Locations in the random layout were randomly generated inside the unit square ($[0, 1] \times [0, 1]$). Locations in the gridded layout were placed on a fixed, equally spaced grid inside the unit square. The two response types were normal and lognormal. For the normal response type, the response was simulated using mean-zero random errors with the exponential covariance (Equation$~$\ref{equation:expcov}) for varying proportions of dependent random error. The proportion of dependent random error is represented by $\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)$, where $\sigma^2_1$ and $\sigma^2_2$ are the dependent random error variance (partial sill) and independent random error variance (nugget), respectively, from Equation$~$\ref{equation:expcov}. The total variance, $\sigma^2_1 + \sigma^2_2$, was always 2. The range was always $\sqrt{2} / 3$, which means that the correlation in the dependent random error decayed to nearly zero at the largest possible distance between two units in the domain. For the lognormal response type, the response was first simulated using the same approach as for the normal response type, except that the total variance was 0.6931 instead of 2. The response was then exponentiated, yielding a random variable whose total variance is 2. The lognormal responses were used to evaluate performance of the sampling-analysis approaches for data that were skewed (i.e., not normal).

```{r, echo = FALSE, results = "asis"}
n <- c("50", "100", "200")
site_locations <- c("Random", "Gridded", "")
psill_ratio <- c("0", "0.5", "0.9")
resp_type <- c("Normal", "Lognormal", "")
parm_tab <- rbind(n, site_locations, psill_ratio, resp_type)
parm_tab[c(2, 4), 3] <- "-"
row.names(parm_tab) <- c("Sample Size (n)", "Location Layout", 
                         "Proportion of Dependent Error", "Response Type")
# parm_tab <- parm_tab[c(1, 2, 4, 3), ]
parm_xtab <- xtable(parm_tab, caption = "\\label{tab:parmtab} Simulation scenario options. All combinations of sample size, location layout, response type, and proportion of dependent random error composed the 36 simulation scenarios. In each simualtion scenario, the total variance was two.")
align(parm_xtab) <- "r|lll"
print(parm_xtab, include.colnames = FALSE, hline.after = c(0, nrow(parm_xtab)), comment = FALSE)
```

@@

The Environmental Protection Agency (EPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) in the United States to assess the water quality of various bodies of water. We will use data from the 2012 National Lakes Assessment (NLA), which measures various aspects of lake health and water quality for lakes in the contiguous United States [@USEPA2012NLA]. Specifically, we will analyze mercury concentration in lakes. Although we know the true mean mercury concentration values for the 986 lakes from the 2012 NLA, we will explore whether or not we obtain an adequately precise estimate for the realized mean mercury concentration if we sample only 100 of the 986 lakes. For each of the four familiar sampling-analysis combinations (IRS-Design, IRS-Model, GRTS-Design, and GRTS-Model), we estimate

@@

```{r apptab, results = "asis"}
res_df <- readr::read_csv(here("inst", "output", "application", "application.csv"))
res_df_drop <- res_df[ ,-2]
names(res_df_drop) <- c("Approach", "Estimate", "SE", "95% LB", "95% UB")
xtab_app <- xtable(res_df_drop, digits = 1, caption = "\\label{tab:appliedtab} Application of design-based and model-based approaches to the NLA data set on mercury concentration. The true mean concentration is 103.2 ng / g.")
print(xtab_app, include.rownames = FALSE, comment = FALSE)
```

>>>>

---
title: A comparison of design-based and model-based approaches for finite population spatial data.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  1.   The design-based and model-based approaches to frequentist statistical inference lie on fundamentally different foundations. In the design-based approach, inference depends on random sampling. In the model-based approach, inference depends on distributional assumptions. We compare the approaches for finite population spatial data.
  2. We provide relevant background for the design-based and model-based approaches and then study their performance using simulations and an analysis of real mercury concentration data. In the simulations, a variety of sample sizes, location layouts, dependence structures, and response types are considered. In the simualtions and real data analysis, the population mean is the parameter of interest and performance is measured using statistics like bias, squared error, and interval coverage.
  3. When studying the simulations and mercury concentration data, we found that regardless of the strength of spatial dependence in the data, sampling plans that incorporate spatial locations (spatially balanced samples) generally outperform sampling plans that ignore spatial locations (non-spatially balanced samples). We also found that model-based approaches tend to outperform design-based approaches, even when the data are skewed (and by consequence, the model-based distributional assumptions violated). The performance gap between the analysis approaches is small when spatially balanced samples are used but large when non-spatially balanced samples are used. This suggests that the sampling choice (whether to select a sample that is spatially balanced) is most important when performing design-based inference.
  4. There are many benefits and drawbacks to the design-based and model-based approaches for finite population spatial data that practitioners must consider when choosing between them. We provide relevant background contextualizing each approach and study their properties in a variety of scenarios, making recommendations for use based on the practitioner's goals.
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \usepackage{caption}
  \usepackage{subcaption}
  \usepackage{setspace}
  \doublespacing
output: rticles::elsevier_article
---

@@

There are two general approaches for using data to make frequentist statistical inferences about a population: design-based and model-based. When data cannot be collected for all units in a population (i.e., population units), data are collected on a subset of the population units. This subset of population units is called a sample. In the design-based approach, inferences about the underlying population are informed via a probabilistic process that randomly assigns some population units to be in the sample. Alternatively, in the model-based approach, inferences are made from specific assumptions about the underlying process generating the data. Each paradigm has a deep historical context [@sterba2009alternative] and its own set of benefits and drawbacks [@hansen1983evaluation].

@@

Though the design-based and model-based approaches apply to statistical inference in a broad sense, we focus on comparing these approaches for spatial data. We define spatial data as data that incorporates the specific locations of the population units into either the design or estimation process. @de1990model give an early comparison of design-based and model-based approaches for spatial data, quashing the belief that design-based approaches could not be used for spatially correlated data. Since then, there have been several general comparisons between design-based and model-based approaches for spatial data [@brus1997random; @verhoef2002sampling; @verhoef2008spatial; @wang2012review; @brus2021statistical]. @cooper2006sampling reviews the two approaches in an ecological context before introducing a "model-assisted" variance estimator that combines aspects from each approach. In addition to @cooper2006sampling, there has been substantial research and development into estimators that use both design and model-based principles (see e.g., @sterba2009alternative and @cicchitelli2012model, and see @chan2020bayesian for a Bayesian approach).

<!-- Paragraph about the contribution of the manuscript and outlining the rest -->
Certainly comparisons between design-based and model-based approaches to spatial data have been studied. But no numerical comparison has been made between design-based approaches that incorporate spatial information and model-based approaches. In this manuscript, we compare design-based approaches that incorporate spatial information to model-based approaches for finite population spatial data. A finite population contains a finite number of population units (we assume the finite number is known); an example is lakes (treated as a whole with the lake centroid representing location) in the contiguous United States. Though we manuscript focuses on finite populations, these comparisons generalize to infinite populations as well. An infinite population contains an infinite number of population units; an example is locations within a single lake. In this manuscript we assume the number of finite population units is known

The rest of the manuscript is organized as follows. In Section \ref{sec:background}, we introduce and provide relevant background for the design-based and model-based approaches to finite population spatial data. In Section \ref{sec:mm}, we describe how we compare performance of the approaches with a simulation study and an analysis of real data that contains mercury concentration in lakes located in the contiguous United States. In Section \ref{sec:results}, we present results from the simulation study and the mercury concentration analysis. And in Section \ref{sec:discussion}, we end with a discussion and provide directions for future research.

@@

The design-based approach assumes the population is fixed. Randomness is incorporated via the selection of units in a sampling frame. A sampling frame is the set of all units available to be sampled. Units from the sampling frame are selected as part of the sample according to a sampling design, which assigns a positive probability of inclusion (inclusion probability) to each unit from the sampling frame. These inclusion probabilities are later used to analyze data. Some examples of commonly used sampling designs include simple random sampling, stratified random sampling, and cluster sampling.

When sampling designs incorporate spatial locations into sampling, we call the resulting samples "spatially balanced." One approach to selecting spatially balanced samples is the Generalized Random Tessellation Stratified (GRTS) algorithm [@stevens2004spatially], which we discuss in more detail in Section \ref{subsec:spb_design}. When sampling designs do not incorporate spatial locations into sampling, we call the resulting samples "non-spatially balanced."

@@

The model-based approach assumes the data are a random realization of a data-generating stochastic process. Randomness is incorporated through distributional assumptions on this process. Strictly speaking, randomness need not be incorporated through random sampling, though @diggle2010geostatistical warn against preferential sampling. Preferential sampling occurs when the process generating the data locations and the process being modeled are not independent of one another. To guard against preferential sampling, model-based approaches often still implement some form of random sampling. When model-based approaches implement random sampling, the inclusion probabilities are ignored when analyzing the data (in contrast to the design-based approach, which relies on these inclusion probabilities to analyze the data).

@@

We previously mentioned that the design-based approach can be used to select spatially balanced samples (samples that incorporate spatial locations of the population units). Spatially balanced samples are useful because parameter estimates from these samples tend to vary less than parameter estimates from samples that are not spatially balanced [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @wang2013design; @benedetti2017spatiallyreview]. The first spatially balanced sampling algorithm to see widespread use was the Generalized Random Tessellation Stratified (GRTS) algorithm [@stevens2004spatially]. To quantify the spatial balance of a sample, @stevens2004spatially proposed loss metrics based on Voronoi polygons (Dirichlet Tessellations). After the GRTS algorithm was developed, several other spatially balanced sampling algorithms emerged, such as the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance Sampling [@benedetti2017spatially], and Halton Iterative Partitioning Sampling [@robertson2018halton]. In this manuscript, we select spatially balanced samples using the Generalized Random Tessellation Stratified (GRTS) algorithm because it has several attractive properties. More specifically, the GRTS algorithm accommodates finite and infinite sampling frames, equal, unequal, and proportional (to size) inclusion probabilities, legacy (historical) sampling [@foster2017spatially], a minimum distance between units in a sample, and replacement units (replacement units are population units that can be sampled when a population unit originally selected can no longer be sampled). The GRTS algorithm selects samples by utilizing a particular mapping between two-dimensional and one-dimensional space that preserves proximity relationships. Via this mapping, units in two-dimensional space are partitioned using a hierarchical address. This hierarchical address is used to map population units to a one-dimensional line. On the one dimensional line, each population unit's line length equals its inclusion probability. Then, a systematic sample of population units is selected on the line and mapped back to two-dimensional space, yielding the desired sample. @stevens2004spatially provide more technical details.

After selecting a sample and collecting data, unbiased estimates of population means and totals can be obtained using the Horvitz-Thompson estimator [@horvitz1952generalization]. If $\tau$ is a population total, the Horvitz-Thompson estimator for $\tau$, denoted by $\hat{\tau}_{ht}$, is is given by
\begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align}
where $Z_i$ is the value of the $i$th population unit in the sample and $\pi_i$ is the inclusion probability of the $i$th population unit in the sample. An estimate of the population mean is obtained by dividing $\hat{\tau}_{ht}$ by $N$, the number of population units.

@@

Finite Population Block Kriging (FPBK) is a model-based approach that expands the geostatistical Kriging framework to the finite population setting [@verhoef2008spatial]. Instead of developing inference based on a specific sampling design, we assume the data are generated by a spatial stochastic process. We summarize some of the basic principles of FBPK next (for more technical details, see @verhoef2008spatial). Let ${\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}$ be an $N \times 1$ response vector at locations $s_1$, $s_2$, . . . , $s_N$ that can be measured at the $N$ population units. Suppose we want to use a sample to predict some linear function of the response variable, $f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}$, where $\mathbf{b}^\prime$ is a $1 \times N$ vector of weights (e.g, the population mean is represented by a weights vector whose elements all equal one). Denoting quantities that are part of the sampled population units with a subscript \emph{s} and quantities that are part of the unsampled population units with subscript \emph{u}, let

@@

FBPK assumes $\bm{\delta}$ in Equation$~$\ref{equation:Zmarginal} has mean-zero and a spatial dependence structure that can be modeled using a covariance function. This covariance function is commonly assumed to be non-negative, second-order stationary (depending only on the distance between population units), isotropic (independent of direction), and decay with distance between population units [@cressie1993statistics]. Henceforth, it is implied that we have made these same assumptions regarding $\bm{\delta}$, though @chiles1999geostatistics, pp. 80-93 discuss covariance functions that are not second-order stationary, not isotropic, or not either. A variety of flexible covariance functions can be used to model $\bm{\delta}$ [@cressie1993statistics]; one example is the exponential covariance function (for a thorough list of spatial covariance functions, see @cressie1993statistics. The $i,j$th element of the exponential covariance matrix, $\cov(\bm{\delta})$, is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align}
where $\sigma^2_{1}$ is the variance parameter quantifying the variability that is dependent (coarse-scale), $\sigma^2_{2}$ is the variance parameter quantifying the variability that is independent (fine-scale), $\phi$ is the range parameter measuring the distance-decay rate of the covariance, and $h_{i,j}$ is the Euclidean distance between population units $i$ and $j$. The proportion of variability attributable to dependent random error is $\sigma^2_{1} / (\sigma^2_{1} + \sigma^2_{2})$. Similarly, the proportion of variability attributable to independent random error is $\sigma^2_{2} / (\sigma^2_{1} + \sigma^2_{2})$. Finally we note that $\sigma^2_{1}$ and $\sigma^2_{2}$ are often called the partial sill and nugget, respectively.

@@

Other approaches, such as k-nearest-neighbors [@fix1989discriminatory; @ver2013comparison] and random forests [@breiman2001random], among others, could also be used to obtain predictions for a mean or total from spatially correlated responses of a finite population. Compared to the k-nearest-neighbors and random forest approach, we prefer FBPK because it is model-based and relies on theoretically-based variance estimators leveraging the model's spatial covariance structure, whereas k-nearest-neighbors and random forests use ad-hoc variance estimators [@ver2013comparison]. Additionally, @ver2013comparison studied compared FBPK, k-nearest-neighbors, and random forests in a variety of spatial data contexts, and FBPK tended to perform best.

@@

We used a simulation study to investigate performance of four sampling-analysis combinations: IRS sampling with a design-based analysis, called "IRS-Design"; IRS sampling with a model-based analysis, called "IRS-Model"; GRTS sampling with a design-based analysis, called "GRTS-Design"; GRTS sampling with a model-based analysis, called "GRTS-Model". These combinations are also provided in Table \ref{tab:designanalysis}.

@@

Performance for the four sampling-analysis combinations was evaluated in 36 different simulation scenarios. The 36 scenarios resulted from the crossing of three sample sizes, two location layouts, two response types, and three proportions of dependent random error. The three sample sizes ($n$) were $n = 50, n = 100,$ and $n = 200$. Samples were always selected from a population size ($N$) of $N = 900$. The two location layouts (of the population units) were random and gridded. Locations in the random layout were randomly generated inside the unit square ($[0, 1] \times [0, 1]$). Locations in the gridded layout were placed on a fixed, equally spaced grid inside the unit square. The two response types were normal and lognormal. For the normal response type, the response was simulated using mean-zero random errors with the exponential covariance (Equation$~$\ref{equation:expcov}) for varying proportions of dependent random error. The proportion of dependent random error is represented by $\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)$, where $\sigma^2_1$ and $\sigma^2_2$ are the dependent random error variance (partial sill) and independent random error variance (nugget), respectively, from Equation$~$\ref{equation:expcov}. The total variance, $\sigma^2_1 + \sigma^2_2$, was always 2. The range was always $\sqrt{2} / 3$, which means that the correlation in the dependent random error decayed to nearly zero at the largest possible distance between two population units in the domain. For the lognormal response type, the response was first simulated using the same approach as for the normal response type, except that the total variance was 0.6931 instead of 2. The response was then exponentiated, yielding a lognormal random variable whose total variance is 2. The lognormal responses were used to evaluate performance of the sampling-analysis approaches for data that were skewed (i.e., not normal).

```{r, echo = FALSE, results = "asis"}
n <- c("50", "100", "200")
site_locations <- c("Random", "Gridded", "")
psill_ratio <- c("0", "0.5", "0.9")
resp_type <- c("Normal", "Lognormal", "")
parm_tab <- rbind(n, site_locations, psill_ratio, resp_type)
parm_tab[c(2, 4), 3] <- "-"
row.names(parm_tab) <- c("Sample Size (n)", "Location Layout", 
                         "Proportion of Dependent Error", "Response Type")
# parm_tab <- parm_tab[c(1, 2, 4, 3), ]
parm_xtab <- xtable(parm_tab, caption = "\\label{tab:parmtab} Simulation scenario options. All combinations of sample size, location layout, response type, and proportion of dependent random error composed the 36 simulation scenarios. In each simualtion scenario, the total variance was 2.")
align(parm_xtab) <- "r|lll"
print(parm_xtab, include.colnames = FALSE, hline.after = c(0, nrow(parm_xtab)), comment = FALSE)
```

@@

The United States Environmental Protection Agency (USEPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) to assess the water quality of various bodies of water in the contiguous United States. We will use data from the 2012 National Lakes Assessment (NLA), which measures various aspects of lake health and water quality [@USEPA2012NLA]. Specifically, we will analyze mercury concentration in lakes. Although we can calculate the true mean mercury concentration values for the 986 lakes from the 2012 NLA, we will explore whether or not we obtain an adequately precise estimate for the realized mean mercury concentration if we sample only 100 of the 986 lakes. For each of the four familiar sampling-analysis combinations (IRS-Design, IRS-Model, GRTS-Design, and GRTS-Model), we estimate (design-based) or predict (model-based) the mean mercury concentration and construct 95% confidence (design-based) or prediction (model-based) intervals from this sample of 100 lakes, which we compare to the actual mean from all 986 lakes.

@@

```{r apptab, results = "asis"}
res_df <- readr::read_csv(here("inst", "output", "application", "application.csv"))
res_df_drop <- res_df[ ,-2]
names(res_df_drop) <- c("Approach", "Est/Pred", "SE", "95% LB", "95% UB")
xtab_app <- xtable(res_df_drop, digits = 1, caption = "\\label{tab:appliedtab} For each sampling-analysis combination (Approach), estimates/predictions (Est/Pred), standard errors (SE), lower 95\\% interval bounds (95\\% LB), and upper 95\\% interval bounds (95\\% UB) for mean mercury concentration computed using the sample of 100 lakes in the NLA data. The true mean concentration of all 986 lakes in the NLA data is 103.2 ng / g.")
print(xtab_app, include.rownames = FALSE, comment = FALSE)
```
